dispRity manual

### 2017-07-31 doc/dispRity-manual.Rmd
Disparity manual updated

```{r, eval = FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

@@

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if Claddis is installed BUILD ONLY
library(dispRity)
```

Note this uses the `release` branch (version`0.4`). <!-- add version !-->
For the piping-hot (but potentially full of bugs) version, you can change the argument `ref = release` to `ref = master`.
`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, and `RCurl`).

@@

This package is not readily available on CRAN.
This is mainly because some parts are still in development and that I prefer the reactivity of GitHub for implementing new suggestions from users.
However, the package follows the strict (and useful!) CRAN standards via Travis. <!-- add link !-->

@@

If you need help about the package, I hope that the following manual will be useful.
If you still need more help, do not hesitate to drop me an email!

Parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

@@

Note that some several options are available, namely which type of distance should be computed.
See more info in the function manual (`?Claddis.ordination`).
Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.

@@

The package can also easily take data from `geomorph` using the `geomorph.ordination` function.
This function simply take procrustes data an performs an ordination:

@@

Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
Additionally you can provide to the function a `geomorph.data.frame` object.
If the latter contains sorting information (i.e. factors), they can be directly used to make a `dispRity` object!

@@

We will see what are these `dispRity` objects in the sections below...

@@

## Ordinating the matrix using cmdscale() with k = 5 dimensions
ordinated_matrix <- cmdscale(eurodist, k = 5)
head(ordinated_matrix)
```
This results in a ordinated matrix with European cities as elements and 5 dimensions.
Note that the number of dimensions should not exceed the number of elements - 1.

Of course any other methods for creating the ordination matrix is totally valid!
The only requirements for `dispRity` is to use matrices with elements as rows and dimensions as columns...

@@

Two functions allow to run templated analysis simply using an ordination matrix: either calculating the disparity through time or the disparity per groups.

@@

For illustrating the following functions, we will use data from Beck and Lee 2015. <!-- ADD PAPER CITE HERE -->
This dataset contains an ordinated matrix of 50 mammals' discrete characters (called `BeckLee_mat50`), a matrix of the same 50 mammals and the estimate discrete data from their descendants (thus 50 + 49 rows, called `BeckLee_mat99`), a data frame containing the ages of each taxa in the dataset (called `BeckLee_ages`) and finally a phylogenetic tree with the relations of the 50 mammals (called `BeckLee_tree`).

```{r, fig.width=7, fig.height=7}
## Setting the random seed for repeatability
set.seed(123)

## Loading the ordinated matrices
data(BeckLee_mat50)
data(BeckLee_mat99)

@@

## The first 5 taxa and dimensions of the 99 taxa + ancestors matrix
head(BeckLee_mat99[,1:5])
tail(BeckLee_mat99[,1:5])

@@

Note of course that you can use your data obtain from the previous chapter!

@@

  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time, you can simply do `my_tree$root.time <- my_age`.
  * A required number of time subsamples (here 3)
  * Your definition of disparity (here the sum of variances)

Note that in the following example, the matrix and the tree comes from the previous sections

@@

When display our disparity data, we can now have information on the operations done to the matrix:

```{r}
## Printing what's in the disparity data
disparity_data
```
Indeed, we asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times and the metric used was the sum of variances.

@@

## Testing for an eventual difference between the time bins
disp_aov <- test.dispRity(disparity_data, test = aov, comparisons = "all")
summary(disp_aov)
```

Please refer to the specific tutorials below for (much!) more information on the nuts and bolts of the package.
You can also directly explore the specific functions manuals within R and navigate to they're related functions.

@@

Similarly as above, we can display information about both groups by simply calling it and then summarising the data (and plotting it!) and performing a statistical test to compare both groups (here a Wilcoxon test).

@@

## Testing for an eventual difference between the groups
test.dispRity(disparity_data, test = wilcox.test, details = TRUE)

```

@@

This following section contains specific information relative to some functions.
If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email. <!-- Add anchor to contacts -->

@@

 *  discrete time subsamples (or time-binning) using `method = discrete`;
 *  continuous time subsamples (or time-slicing) using `method = continuous`.

For the time-slicing method details see @GuillermeSTD.
<!-- TODO: link to the time-slicing algorithm explained! -->
For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:

 *  defining the boundaries of the time bins (when `method = discrete`);
 *  defining the time slices (when `method = continuous`).

@@

```{r, eval=FALSE}
## Generating three time bins containing the data present every 40 Ma
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = c(120, 80, 40, 0))
```

In this example, the taxa were split inside each time-bin according to their age.
However, the taxa here are considered like single points in time.
It is totally possible that the taxa in question could have a longer longevity (e.g. if the taxonomic focus is at the genera level or above) and that it spans across multiple time bins.
In this case, it is possible to include them in more than one bin by providing a table of First and last occurrence dates (FAD/LAD).
This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:

```{r, eval=FALSE}
## Displaying the table of first and last occurrence dates for each taxa
head(BeckLee_ages)

@@

```{r, eval=FALSE}
## Automatically generate three equal length bins:
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = 3)
```

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age + 1% of the tree length as long as at least three elements are present at that point in time.
The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.

@@

```{r, eval=FALSE}
## Generating four time slices every 40 million years under gradual evolution
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Generating four time slices automatically
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
```

@@

One important step in analysing ordinated matrices is to pseudo-replicate the data to seem how robust the results of future analysis are sensitive to outliers in the dataset.
This can be achieved using the function `boot.matrix` to bootstrap or rarefy the ordinated matrix.
The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):

```{r, eval=FALSE}
## Default bootstrapping
boot.matrix(data = BeckLee_mat50)
```

@@

 *  `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any elements drawn from the dat);
 *  `single` where only one random element is replaced by one other random element for each pseudo-replicate.

```{r, eval=FALSE}
## Bootstrapping with the single bootstrap method
boot.matrix(BeckLee_mat50, boot.type = "single")
```

This function also allows to rarefy the data using the `rarefaction` argument.
Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
For example, one might be interested in looking at the effect of reducing the number of elements to the results of the analysis.

@@

```{r, eval=FALSE}
## Bootstrapping with the full rarefaction
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)

@@

```{r, eval = FALSE}
## Removing the last axis 50% last dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

@@

There are many ways for measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space
<!-- Cite Donohue -->
, the sum and the product of the ranges and variances
<!-- Cite Wills -->
 or the median position of the elements relative to their centroid.
 <!-- Cite Sive -->
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.

@@

Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix or the `ranges` one that calculates the range of each dimension.

@@

## Measuring disparity as the standard deviation of the variance of each axis of
## the variance covariance matrix (level 1, level 2 and level 3 functions).

summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

@@

#### What level is the disparity?

In these examples, we considered disparity to be a single value.
For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
This results in single value to be calculated.

In some cases however, it might also be interesting to calculate disparity as a distribution.
For example, one of the metrics commonly used in palaeobiology is the sum of variances of each dimensions (`metric = c(sum, variances)`) [@Wills1994].
However, it might be interesting to look directly at the distribution of the variances itself.

```{r}
## A one dimensional disparity value
(sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(median, variances))))

## A two dimensional disparity
(summary(all_variances <- (dispRity(BeckLee_mat50, metric = variances))))
```

Looking at all the variances rather than simply their median can tell us that (by the look of it) the variances of each dimensions seems to be normally distributed.

> It is also possible to reuse an already calculated disparity distribution and summarise it using a level 1 metric:

```{r}
## Transforming the two dimensional one in a one dimensional one
sum_variances2 <- summary(dispRity(all_variances, metric = median))

## Both ways output the same disparity values:
sum_variances1 == sum_variances2
```

@@

1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).

@@

### Metrics run through
Here is a run through of the main metrics implemented in `dispRity`.
First, we will create a dummy ordinated space using the `space.maker` utility function:

@@

## Calculating the volume with different options to be passed to hypervolume::hypervolume
summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100, quantile = 0.5))
```
> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [Curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) which often results in near 0 disparity values.

@@

> This metric is only mathematically valid if the dimensions within the space are all orthogonal!

@@

```{r}
## Calculating the distribution of the distances between each elements and the centroid
summary(dispRity(dummy_space, metric = centroids))

## Calculating disparity as the median value of these distances
summary(dispRity(dummy_space, metric = c(median, centroids)))
```

@@

```{r}
## Calculating the distance between the elements and the origin of the ordinated space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))

## Calculating disparity as the distance between the elements and a specific point
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
```

@@

Because of its architecture, printing `dispRity` objects only summarise their content but does not print the disparity value measured or associated analysis.
<!-- link to dispRity objects -->
To actually summary what's in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphic representation of the results.

@@

The summary functions can also take various options such as the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated); the `cent.tend` for the  central tendency to use for summarising the results (default is `median`); the `rounding` option corresponding to the number of decimals to print (default is `2`) and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)
These options can easily be changed from the defaults as follows:

@@

```{r, eval = FALSE}
## Loading the xtable package
require(xtable)
## Converting the table in LaTeX
xtable(summary(disparity_time_slices))
```

@@

The plots can be of four different types: `continuous` for displaying continuous disparity curves and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.

@@

 *  `pairwise` (default): to compare each subsamples pairwise

 *  `referential`: to compare each subsamples to the first one

 *  `sequential`: to compare each subsamples to the following one

 *  `all`: to compare all the subsamples together (like in analysis of variance)

@@

```{r}
## Performing a t-test to the test a difference in disparity between crown and stem mammals
test.dispRity(disparity_crown_stem, test = t.test)

@@

## Performing a Wilcoxon test applied to sliced disparity the data sequentially with correction
test.dispRity(disparity_time_slices, test = wilcox.test, comparisons = "sequential",
    correction = "bonferroni")

## Measuring the overlap between distributions in the time bins (using the implemented
## Bhattacharyya Coefficient function - see ?bhatt.coeff)
test.dispRity(disparity_time_bins, test = bhatt.coeff)
```

@@

```{r}
## Performing and ANOVA applied to the same data
slice_aov <- test.dispRity(disparity_time_slices, test = aov, comparisons = "all")
slice_aov

@@

## That can be summarised using summary
summary(slice_aov)
```
Of course, following the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
The only condition is that the test can be applied to at least two distributions.
In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `aov` for example).

@@

Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
This value can be of course pseudo-replicated (boostraped) in order to obtain a more confident distribution of this summary metric.
However, another way to perform disparity analysis is to use the whole distribution rather than their summary (e.g. the variances or the ranges).

@@

```{r, fig.width=12, fig.height=6, eval = FALSE}
## Summary table of the data
summary(disparity)

@@

>>>>

```{r, eval=FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

@@

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if Claddis is installed BUILD ONLY
library(dispRity)
## Setting the random seed for repeatability
set.seed(123)
```

Note this uses the `release` branch (version`0.4`). <!-- @@@ -->
For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).

@@

This package is not readily available on CRAN.
This is mainly because some parts are still in development and that the reactivity of GitHub is better for implementing new suggestions from users.
However, the package follows the strict (and useful!) CRAN standards via Travis.

[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)

@@

If you need help about the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

@@

Note that several options are available, namely which type of distance should be computed.
See more info in the function manual (`?Claddis.ordination`).
Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.

@@

You can also easily use data from `geomorph` using the `geomorph.ordination` function.
This function simply take procrustes data and performs an ordination:

@@

Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
Additionally you can provide to the function a `geomorph.data.frame` object.
If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!

@@

More about these `dispRity` objects below!

@@

## Ordinating the matrix using cmdscale() with k = 5 dimensions
ordinated_matrix <- cmdscale(eurodist, k = 5)
head(ordinated_matrix)
```

  This results in a ordinated matrix with European cities as elements and 5 dimensions.

Of course any other methods for creating the ordination matrix is totally valid - or not use any ordination at all!
The only requirements for the `dispRity` functions is to use matrices with elements as rows and dimensions as columns...

@@

Two functions allow to run pipelined analysis simply using an ordination matrix: either calculating the disparity through time or the disparity per groups.

@@

For illustrating the following functions, we will use data from @beckancient2014.
This dataset contains an ordinated matrix from 50 mammals' discrete characters (called `BeckLee_mat50`), another matrix of the same 50 mammals and the estimate discrete data from their descendants (thus 50 + 49 rows, called `BeckLee_mat99`), a data frame containing the ages of each taxa in the dataset (called `BeckLee_ages`) and finally a phylogenetic tree with the relations between the 50 mammals (called `BeckLee_tree`).

```{r, fig.width=7, fig.height=7}
## Loading the ordinated matrices
data(BeckLee_mat50)
data(BeckLee_mat99)

@@

## The first 5 taxa and dimensions of the 99 taxa + ancestors matrix
BeckLee_mat99[c(1,2,98,99),1:5]

@@

Of course you can use your own data as detailed in the [previous chapter](#which-data).

@@

  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time, you can simply do `my_tree$root.time <- my_age`.
  * A required number of time subsamples (here 3)
  * Your very own definition of disparity (here the sum of variances)

In the following example, the matrix and the tree comes from the [previous sections](#example-data).

@@

This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
When displayed these `dispRity` objects provides us we information on the operations done to the matrix:

```{r}
## Printing what's in the disparity data
disparity_data
```

Indeed, we asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.

@@

## Testing for an eventual difference between the time bins
disp_aov <- test.dispRity(disparity_data, test = aov, comparisons = "all")
summary(disp_aov)

## Note that anova assumptions where not checked here!
```

Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
You can also directly explore the specific functions manuals within R and navigate to they're related functions.

@@

We can display information about both groups by simply calling the output variable and then summarising the data (and plotting it!) or/and performing a statistical test to compare both groups (here a Wilcoxon test).

@@

## Testing for an eventual difference between the groups
test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
```

@@

This following section contains specific information relative to some functions.
If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email (<guillert@tcd.ie>).
The several tutorials below describe specific functionalities of some specific functions, please always refer to the function manuals for the full function documentation!

@@

 *  discrete time subsamples (or time-binning) using `method = discrete`
 *  continuous time subsamples (or time-slicing) using `method = continuous`

For the time-slicing method details see @GuillermeSTD.
For both methods, the function intakes the `time` argument which can be a vector of `numeric` values for:

 *  defining the boundaries of the time bins (when `method = discrete`)
 *  defining the time slices (when `method = continuous`)

@@

```{r, eval=TRUE}
## Generating three time bins containing the data present every 40 Ma
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = c(120, 80, 40, 0))
```

```{r, eval=TRUE}
## Automatically generate three equal length bins:
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = 3)
```

In this example, the taxa were split inside each time-bin according to their age.
However, the taxa here are considered as single points in time.
It is totally possible that some taxa could have had longer longevity and that it spans across multiple time bins.
In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:

```{r, eval=TRUE}
## Displaying the table of first and last occurrence dates for each taxa
head(BeckLee_ages)

@@

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as being the root age plus 1% of the tree length as long as at least three elements are present at that point in time.
The algorithm adds an extra 1% tree length until reaching the required minimum of three elements.
It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance between tips *and* nodes.

@@

```{r, eval=TRUE}
## Generating four time slices every 40 million years under gradual evolution
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)

## Generating four time slices automatically
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
```

## Customised subsamples

Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-between-groups).
This function simply takes the list of elements to put in each group (whether they are the actual element names or their idea).

```{r, eval=TRUE}
## Creating the two groups as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44))

## Separating the dataset into two different groups
custom.subsamples(BeckLee_mat50, group = mammal_groups)
```

Elements can easily be assigned to different groups if necessary!

```{r, eval=FALSE}
## Creating the three groups as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44).
                      "all" = c(1:50))
```

@@

One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results of future analysis are sensitive to outliers in the dataset.
This can be achieved using the function `boot.matrix` to bootstrap or/and rarefy the data.
The default options will bootstrapping the matrix 100 times without rarefaction using the "full" bootstrap method (see below):

```{r, eval=TRUE}
## Default bootstrapping
boot.matrix(data = BeckLee_mat50)
```

@@

 * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
 * `single` where only one random element is replaced by one other random element for each pseudo-replicate

```{r, eval=TRUE}
## Bootstrapping with the single bootstrap method
boot.matrix(BeckLee_mat50, boot.type = "single")
```

This function also allows to rarefy the data using the `rarefaction` argument.
Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
For example, one might be interested in looking at the effect of reducing the number of elements on the results of an analysis.

@@

```{r, eval=TRUE}
## Bootstrapping with the full rarefaction
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)

@@

```{r, eval=TRUE}
## Removing the last axis 50% last dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

@@

There are many ways for measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @finlay2015morphological.
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.

@@

Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculates the variance or the range of each dimension of the ordinated matrix respectively.

@@

## Measuring disparity as the standard deviation of the variance of each axis of
## the variance covariance matrix (level 1, level 2 and level 3 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

@@

In these examples, we considered disparity to be a single value.
For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
It is however possible to calculate [disparity as a distribution](#disparity-as-a-distribution).

@@

1: This function uses an estimation of the eigen value that only works for MDS or PCoA ordinations (not PCA).

@@

### Metrics run through
Here is a run through of the main metrics implemented in `dispRity`.
First, we will create a dummy ordinated space using the `space.maker` utility function (more about that [here](#space.maker):

@@

## The volume with different options to be passed to hypervolume::hypervolume
summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100,
                 quantile = 0.5))
```
> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [Curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) which often results in near 0 disparity values.

@@

> This metric is only an euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!

@@

```{r}
## The distribution of the distances between each elements and the centroid
summary(dispRity(dummy_space, metric = centroids))

## Disparity as the median value of these distances
summary(dispRity(dummy_space, metric = c(median, centroids)))
```

@@

```{r}
## The distance between the elements and the origin of the ordinated space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))

## Disparity as the distance between the elements and a specific point in space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
```

@@

Because of its architecture, printing `dispRity` objects only summarise their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
To actually summary what's in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphic representation of the results.

@@

The summary functions can also take various options such as:

 * the `quantile` values for the confidence intervals levels (by default, the 50 and 95 quantiles are calculated)
 * the `cent.tend` for the  central tendency to use for summarising the results (default is `median`)
 * the `rounding` option corresponding to the number of decimals to print (default is `2`)
 *  and the `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)

These options can easily be changed from the defaults as follows:

@@

```{r, eval=FALSE}
## Loading the xtable package
require(xtable)
## Converting the table in LaTeX
xtable(summary(disparity_time_slices))
```

@@

The plots can be of four different types:

 * `continuous` for displaying continuous disparity curves
 * and `box`, `lines` and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines and confidence interval polygons.

@@

 * `pairwise` (default): to compare each subsamples pairwise
 * `referential`: to compare each subsamples to the first one
 * `sequential`: to compare each subsamples to the following one
 * `all`: to compare all the subsamples together (like in analysis of variance)

@@

```{r}
## T-test to the test a difference in disparity between crown and stem mammals
test.dispRity(disparity_crown_stem, test = t.test)

@@

## Wilcoxon test applied to sliced disparity the data sequentially with correction
test.dispRity(disparity_time_slices, test = wilcox.test,
              comparisons = "sequential", correction = "bonferroni")

## Measuring the overlap between distributions in the time bins (using the
## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff)
test.dispRity(disparity_time_bins, test = bhatt.coeff)
```

@@

```{r}
## Performing and ANOVA applied to the same data
(slice_aov <- test.dispRity(disparity_time_slices, test = aov,
                            comparisons = "all"))

@@

## That can be summarised using summary
summary(slice_aov)
```

Of course, following the modular design of the package, tests can always be made by the user (the same way disparity metrics can be user made).
The only condition is that the test can be applied to at least two distributions.
In practice, the `test.dispRity` function will pass the calculated disparity data (distributions) to the provided function in either pairs of distributions (if the `comparisons` argument is set to `pairwise`, `referential` or `sequential`) or a table containing all the distributions (`comparisons = all`; this should be in the same format as data passed to `aov` for example).

@@

Disparity is often regarded as a summary value of the position of the all elements in the ordinated space.
For example, the sum of variances, the product of ranges or the median distance between the elements and their centroid will summarise disparity as a single value.
This value can be of course pseudo-replicated (bootstrapped) in order to obtain a more confident distribution of this summary metric.
However, another way to perform disparity analysis is to use the whole distribution rather than their summary (e.g. the variances or the ranges).

@@

```{r, fig.width=12, fig.height=6, eval=TRUE}
## Summary table of the data
summary(disparity)

@@

### `space.maker`
Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
This function allows to simulated multidimensional spaces given a certain number of properties.
It takes as arguments the number of elements (data points) and dimensions to create the space but also allows more fine tuning data simulation: it is possible to give a specific distribution to each dimensions, provide a correlation matrix to link the dimensions or even providing an *a priori* distribution of the variance per distributions!

```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(mfrow = (c(2,2)), bty = "n")
## Visualising 3D spaces
require(scatterplot3d)

## A cube space
scatterplot3d(space.maker(2500, 3, runif), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform cube space")

## A plane space
scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform plane space")

## An ellipsoid space (=a spheric space with correlation)
cor_matrix <- matrix(cbind(1,0.8,0.2, 0.8,1,0.7, 0.2,0.7,1), nrow = 3)
scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal ellipsoid space")

## A cylindrical space with a decreasing variance per axis
scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
              scree = c(0.7, 0.2, 0.1)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal cylindrical space")

## Resetting the graphic parameters
par(op)
```
